---
title: "The effect of AMU therapeutic use"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{css horizontal scrolling, echo = FALSE}
pre, code {white-space:pre !important; overflow-x:scroll !important}

html { overflow-x: scroll; }
```

```{r general options, include = FALSE}
knitr::knit_hooks$set(
  margin = function(before, options, envir) {
    if (before) par(mgp = c(1.5, .5, 0), bty = "n", plt = c(.105, .97, .13, .97))
    else NULL
  },
  prompt = function(before, options, envir) {
    options(prompt = if (options$engine %in% c("sh", "bash")) "$ " else "> ")
  })

knitr::opts_chunk$set(cache = FALSE, autodep = TRUE, message = FALSE, warning = FALSE,
                      dev.args = list(pointsize = 11), fig.height = 3.5,
                      fig.width = 4.24725, fig.retina = 2, fig.align = "center")

options(width = 137)
```

## Packages

Installing the required packages:

```{r}
required <- c("dplyr", "magrittr", "purrr", "readr")
to_install <- setdiff(required, row.names(installed.packages()))
if (length(to_install)) install.packages(to_install)
```

Loading `magrittr` for interactive use:

```{r}
library(magrittr)
```

## Loading the data

```{r eval = FALSE}
viparc <- readr::read_csv("https://raw.githubusercontent.com/viparc/prophylactic/master/data/viparc_qualitative.csv",
                          col_types = paste(c("ciii", rep("l", 51)), collapse = ""))
```

```{r eval = TRUE, include = FALSE}
viparc <- readr::read_csv("data/viparc_qualitative.csv",
                          col_types = paste(c("ciii", rep("l", 51)), collapse = ""))
```

We have 3 types of variables in this data frame:

* week ID with the variables `farm`, `flock` and `week`;
* presence of clinical signs with the variables `respiratory`, `diarrhoea`,
  `cns`, `malaise`, `leg_lesions` and `sudden_death`;
* AMU (presence / absence too) with the `r ncol(viparc) - 9` variables
  `amoxicillin_use` to `unknown_use`.

For the purpose of the examples used below, we will used the following sets of
antimicrobials and clinical signs:

```{r}
am_set <- c("colistin_use", "oxytetracycline_use")
cs_set <- c("malaise", "diarrhoea")
gap <- 1
```

## Preparing the data

The function `fill_gaps()` takes a vector of logicals (or zeros and ones) and
replaces all the `1 0 1` sequences by `1 1 1` (when `gap = 1`), or all `1 0 0 1`
sequences by `1 1 1 1` (when `gap = 2`), etc...:

```{r}
fill_gaps_0 <- function(x, gap) {
  as.integer(x) %>%         # because easier to deal with 1-character encoding
    paste(collapse = "") %>% 
    gsub(paste(c(1, rep(0, gap), 1), collapse = ""),
         paste(rep(1, gap + 2), collapse = ""), .) %>% 
    gsub("NA", "x", .) %>%  # need to replace potential NA by 1-character...
    strsplit("") %>%        # ... because of the strsplit() call that follows
    unlist(FALSE) %>%       # because strsplit() returns a list
    sub("x", NA, .) %>%     # back to NA
    as.numeric() %>% 
    as.logical()
}
```

Let's try it:

```{r}
fill_gaps_0(c(TRUE, TRUE, FALSE, FALSE, TRUE, FALSE, FALSE, TRUE, TRUE, TRUE, FALSE, FALSE), 2)
```

Works fine but it doesn't replace the gaps that are shorter than the `gap`
argument:

```{r}
fill_gaps_0(c(TRUE, TRUE, FALSE, TRUE, TRUE, FALSE, FALSE, TRUE, TRUE, TRUE, FALSE, FALSE), 2)
```

`fill_gaps()` is a wrapper around `fill_gaps_0()` that fixes this problem:

```{r}
fill_gaps <- function(x, gap) {
  while(gap > 0) {
    x <- fill_gaps_0(x, gap)
    gap <- gap - 1
  }
  x
}
```

Let's try it:

```{r}
fill_gaps(c(TRUE, TRUE, FALSE, FALSE, TRUE, FALSE, FALSE, TRUE, TRUE, TRUE, FALSE, FALSE), 2)
```

and:

```{r}
fill_gaps(c(TRUE, TRUE, FALSE, TRUE, TRUE, FALSE, FALSE, TRUE, TRUE, TRUE, FALSE, FALSE), 2)
```

The function `episodes()` identifies the different disease episodes. It would
for example transform `1 1 1 0 0 1 1 0 0 0 1 0` into `1 1 1 0 0 2 2 0 0 0 3 0`:

```{r}
episodes <- function(x) {
  x <- rle(x)
  sel <- which(x$values > 0)
  x$values[sel] <- seq_along(sel)
  inverse.rle(x)
}
```

Let's try it:

```{r}
episodes(c(1, 1, 1, 0, 0, 1, 1, 0, 0, 0, 1, 0))
```

The function `split_by_disease_episode()` uses the function `episodes()` to
generate, **for a given flock**, disease episodes according to a set of clinical
signs `cs`, and splits by row the data frame into a list of data frames, each
corresponding to a disease episode:

```{r}
split_by_disease_episode <- function(df, cs, gap) {
  if (any(unique(diff(df$week))) != 1) return(NULL) # makes sure there is no missing week
  df %>% 
    dplyr::mutate(disease = dplyr::select(., !! cs) %>%
                              rowSums() %>%
                              as.logical() %>%
                              fill_gaps(gap) %>% 
                              as.integer() %>%
                              episodes()) %>% 
    dplyr::filter(disease > 0) %>% 
    split(., .$disease) %>% 
    lapply(dplyr::select, -disease)
}
```

Let's try it:

```{r}
viparc %>% 
  dplyr::filter(farm == "75-004", flock == 1) %>% 
  dplyr::arrange(week) %>% 
  split_by_disease_episode(cs_set, gap) %>% 
  lapply(dplyr::select, farm, flock, week, !! cs_set) # for simpler printing
```

Works for one given cycle. Here is how we make it works for all cycles:

```{r}
a <- viparc %>% 
  dplyr::arrange(week) %>% 
  dplyr::group_by(farm, flock) %>% 
  dplyr::group_map(~ split_by_disease_episode(.x, cs_set, gap), keep = TRUE) %>% 
  unlist(FALSE)
```

which gives:

```{r}
a %>% 
  lapply(dplyr::select, farm, flock, week, !! cs_set) %>%
  head()
```

Before we continue, we want to exclude from this list all the disease episodes
with AMU during the $x$ weeks before the start of the episode, AMU being, again,
defined by a set of antimicrobials.

```{r}
exclude_if_amu_before <- function(df, x, df_ref, ab) {
  require(magrittr)
  first_week <- min(df$week)
  if (first_week == 1) return(df)
  weeks <- (first_week - x):(first_week - 1)
  weeks <- weeks[weeks > 0]
  amu_before_episode <- df_ref %>% 
    dplyr::filter(farm == df$farm[1], flock == df$flock[1], week %in% weeks) %>% 
    dplyr::select(ab) %>% 
    rowSums() %>% 
    as.logical() %>% 
    any()
  if (amu_before_episode) return(NULL)
  df
}
```

where `df` is a sub data frame of the `viparc` data frame, `x` is the duration,
in weeks, of the period before the start of the disease episode we consider to
filter out according to AMU, `df_ref` is typically the `viparc` data frame, and
`ab` is a set of antimicrobials used to define AMU. Let's try it:

```{r}
exclude_if_amu_before(a[[1]], 2, viparc, am_set)
```

The function `before_after()` below splits the data frame of a disease episode
into 2 data frames, one before the use of AMU and one after the use of AMU. The
week where AMU starts is deleted. AMU is defined in terms of a set of
antimicrobials.

```{r}
before_after <- function(df, ab) {
  require(magrittr)
  start <- df %>%
    dplyr::select(ab) %>%
    rowSums() %>%
    as.logical() %>%
    which() %>%
    head(1)
  if (length(start) < 1) return(list(df, NULL)) # there is never AMU
  if (start < 3) return(list(NULL, df[-(1:start), ])) # AMU starts on any of the first 2 weeks
  nr <- nrow(df)
  if (start > nr - 2) return(list(df[-(start:nr), ], NULL)) # AMU starts on any of the last 2 weeks
  list(df[1:(start - 1)], df[(start + 1):nr, ])
}
```

Note that we need at least 2 weeks to be able to calculate a mortality rate.
That's why we consider the cases where AMU starts during the first two weeks or
the last two weeks of the episode. Let's try it:

```{r}
b <- lapply(a, before_after, am_set)
```

Finally, since the mortality rate can be calculated only if we have at least 2
consecutive weeks, we need to exclude all the disease episodes that have only 1
week.

```{r}
remove_short_episodes <- function(lst) {
  lst[sapply(lst, nrow) > 1]
} 
```

A pipeline that assembles the above defined `split_by_disease_episode()`,
`exclude_if_amu_before()`, `before_after()` and `remove_short_episdoes()`
functions:

```{r}
c <- viparc %>% 
  dplyr::arrange(week) %>% 
  dplyr::group_by(farm, flock) %>% 
  dplyr::group_map(~ split_by_disease_episode(.x, cs_set, gap), keep = TRUE) %>% 
  unlist(FALSE) %>% 
  lapply(exclude_if_amu_before, 2, viparc, am_set) %>% 
  purrr::compact() %>% # because exclude_if_amu_before() potentially generates NULL elements
  lapply(before_after, am_set) %>% 
  purrr::transpose() %>%
  lapply(purrr::compact) %>%  # because before_after() potentially generates NULL elements
  lapply(remove_short_episodes) %>% 
  setNames(c("before", "after"))
```

Which here gives:

```{r}
sapply(c, length)
```

## Computing mortality rates

The mortality rate is computed

